---
title: Reproduciable Research Peer Assignment 2   
author: "Salim Elzwawi"   
date: July 27, 2014   
---

# Impact of severe weather on the economy and public health in the united states

## Synopsis
This report presents an analysis of the impact of severe weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. Estimates of fatalities, injuries, property and crop damage are used to decide which types of events are most harmful to the population health and economy. The analysis shows that excessive heat and tornados have the highest impact on population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Basic settings
```{r}
echo = TRUE  # make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
```


## Data processing
This section of code is used to download, unzip and read the data. 

```{r download, message=FALSE, cache=TRUE, warning=FALSE}
if(!file.exists("data.csv")) {
  # Download
  url<-"http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url,"data.bz2")
  
  # Unzip
  zz <- readLines(gzfile("data.bz2"))
  zz <- iconv(zz, "latin1", "ASCII", sub="")
  writeLines(zz, "data.csv")
  rm(zz)
  }
 
  # Read data
  Data<-read.csv("data.csv", sep=",", quote = "\"", header=TRUE)
  dim(Data)
  head(Data, n = 2)
```

There are 902297 rows and 37 columns in total.
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, while records of recent years are complete. Therefore, only recent years are considered in the analysis.

```{r}
if (dim(Data)[2] == 37) {
    Data$year <- as.numeric(format(as.Date(Data$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(Data$year, breaks = 30)
```  

The above histogram shows that the number of events recorded starts to increase significantly from 1995. Accordingly, the subset of the data from 1990 to 2011 is considered in the analysis. 
```{r}
storm <- Data[Data$year >= 1995, ]
dim(storm)
```
Now, there are 681500 rows and 38 columns in total.

#### Impact on health
In this section, the number of **fatalities** and **injuries** that are caused by the severe weather events are studied. The 15 most severe types of weather events are looked at. 

```{r}
sortHelper <- function(fieldName, top = 15, dataset = Data) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortHelper("FATALITIES", dataset = storm)
injuries <- sortHelper("INJURIES", dataset = storm)
```

#### Impact on economy
To study the impact on economy, the variables **property damage** and **crop damage** are chosen for the analysis. The data is converted first into comparable numerical forms according to the meaning of units described in the code book ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}
convertHelper <- function(dataset = storm, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}

storm <- convertHelper(storm, "PROPDMGEXP", "propertyDamage")
storm <- convertHelper(storm, "CROPDMGEXP", "cropDamage")
names(storm)
options(scipen=999)
property <- sortHelper("propertyDamage", dataset = storm)
crop <- sortHelper("cropDamage", dataset = storm)
```

### Results
The lists below present the impact of storms on health in terms of fatalities and injuries. 
```{r}
fatalities
injuries
```
The following graphs depict the total fatalities and injuries caused by each of these severe weather events. 
```{r}
fatalitiesPlot <- qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 90, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011") +geom_bar(fill="#FF9999", colour="black")
injuriesPlot <- qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 90, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011") + geom_bar(fill="#FF9999", colour="black")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```  

The above histograms show that **excessive heat** and **tornados** cause most fatalities while **tornados** cause most injuries in the United States from 1995 to 2011.

The two sorted lists below show the impact on the economy in terms of cost of damage.  

```{r}
property
crop
```
The following graphs depict the total property damage and total crop damage caused by these severe weather events. 
```{r}
propertyPlot <- qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1995 - 2011") + geom_bar(fill="#FF9999", colour="black")

cropPlot<- qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1995 - 2011") + geom_bar(fill="#FF9999", colour="black")
grid.arrange(propertyPlot, cropPlot, ncol = 2)
```  

The above histograms show that **flood** and **hurricane/typhoon** cause most property damage while **drought** and **flood** cause most crop damage in the United States from 1995 to 2011.

### Conclusion  
The analysis of the data from 1995 to 2011 reveals that **excessive heat** and **tornado** have the highest impact on population health, while **flood**, **drought** and **hurricane/typhoon** have the greatest economic consequences.



